---
title: "final"
author: "Iris"
date: "December 16, 2014"
output: html_document
---

I created a set of functions that take in starting parameters for a Lotka-Volterra model of two microbes growing and competing in one host, and return the expected average population size of the two microbes after a user-identified number of host generations. Each microbe has an intrinsic rate of increase (the rate at which they would breed absent of any resource competition), a carrying capacity within the host, and a degree to which the population of the other microbe affects their growth rate. I also allow the user to set the number of hosts available to the microbe, and the number of generations the microbes undergo per host generation. 

The function is:

microbe population at time t = population at t-1 * exp(intrinsic growth rate * (1-(population at t - competition term * population of other microbe at time t)))

At low population levels of both microbes, the competition term will be low, and the population will grow at close to its fundamental rate of increase. At higher population levels, competition will reduce the rate to growth to zero. If the population excedes its carrying capacity, it will reduce in size. 

```{r, echo=FALSE, eval=TRUE}
n.host <- 100
hostMat <- matrix(NA, nrow=n.host, ncol=2)


hostMat[,1] <- runif(100, 0, 1)

hostMat[,2] <- 1-hostMat[,1]

n.mic.gen <- 10

# number of offspring per generation

nOff1 <- 2
nOff2 <- 20

# starting population size

micStart <- 1e3
micK <- 1e5

# interaction term
alpha12 <- 0.5
alpha21 <- 0.05

internalLV.show <- function(vector){
  
  tmp <- matrix(NA, nrow=10, ncol=2)
  tmp[1,1] <- vector[1]*micStart
  tmp[1,2] <- vector[2]*micStart
  
  
  for(i in 2:10){
    tmp[i,1] <-tmp[i-1,1]*(exp(nOff1*(1-(tmp[i-1,1]+alpha12*tmp[i-1,2])/micK)))
    
    tmp[i,2] <-tmp[i-1,2]*(exp(nOff1*(1-(tmp[i-1,2]+alpha21*tmp[i-1,1])/micK)))
  }
  return(tmp)
}

show <- internalLV.show(c(50, 50))

plot(x=1:10, y=seq(0, 130000, 130000/9), type="n", xlab="time step", ylab="population")
lines(show[,1], col="blue")
lines(show[,2], col="red")

```


Due to competition between the organisms, their population levels show fluctuation over time within a host. In order to model transimission between hosts, I used a second function with two new parameters, which determine the percentage of each microbe's population that is transmitted to its offspring. I model the population with each host having exactly one offspring. Each new host starts out with excatly 1000 individual microbes, with the makeup of that seed population determined by the microbe's relative frequency in the parent multiplied by that microbe's transmission probability.

```{r, echo=FALSE, eval=TRUE}

nextSeed <- function(oneGenVec, tProb1, tProb2){
  oneTrans <- oneGenVec[1]*tProb1*micStart/((oneGenVec[1]*tProb1+ oneGenVec[2]*tProb2)+1)
  
  twoTrans <- oneGenVec[2]*tProb2*micStart/((oneGenVec[1]*tProb1+ oneGenVec[2]*tProb2)+1)
  
  return(c(oneTrans, twoTrans))
}

```


To make the overall function as quick and user-friendly as possible, I created a function that takes in the user's desired range of competition coeffiecients (ranging from 0 to 1), transmission probabilities (also from zero to 1), and intrinsic rates of growth (unbounded, whole numbers), and makes a matrix of each unique combination of parameters. 

```{r, echo=FALSE, eval=TRUE}
make.input <- function(alphamin, alphamax, tProbmin, tProbmax, nOffmin, nOffmax, step.size){
  alpha <- seq(alphamin, alphamax, step.size)
  tprob <- seq(tProbmin, tProbmax, step.size)
  nOff <- seq(nOffmin, nOffmax)
  alphmat <- cbind(rep(alpha, each=length(alpha)), rep(alpha, length(alpha)))
  
  probmat <- cbind(rep(alphmat[,1], length(tprob)), rep(alphmat[,2], length(tprob)), 
                 rep(tprob, each=nrow(alphmat)))
  
 off1mat <- cbind(rep(probmat[,1], length(nOff)), rep(probmat[,2], length(nOff)), rep(probmat[,3], length(nOff)),
                 rep(nOff, each=nrow(probmat)))
 off2mat <- cbind(rep(off1mat[,1], length(nOff)), rep(off1mat[,2], length(nOff)), 
                rep(off1mat[,3], length(nOff)), rep(off1mat[,4], length(nOff)), rep(nOff, nrow(off1mat)))
 off2mat <- unique(off2mat)
 return(off2mat)
}
```


Finally, I combine the within-host and transmission functions into one large model function, and apply that function to each user-defined parameter combination. The function returns the average population level of each microbe at each parameter combination over N host generations.

```{r, echo=FALSE, eval=TRUE}
inMat <- make.input(0.1, 0.5, 0.2, 0.7, 1, 4, 0.1)
n.gen=10

paramTest <- function(inVec, hostMat){
    
oneMat <- matrix(NA, nrow=n.host, ncol=n.gen)
oneMat[,1] <- hostMat[,1]*micStart

twoMat <- matrix(NA, nrow=n.host, ncol=n.gen)
twoMat[,1] <- hostMat[,2]*micStart

internalLV <- function(vec){
  
  tmp <- matrix(NA, nrow=10, ncol=2)
  tmp[1,1] <- vec[1]
  tmp[1,2] <- vec[2]
  

  for(i in 2:10){
    tmp[i,1] <-tmp[i-1,1]*(exp(inVec[3]*(1-(tmp[i-1,1]+inVec[1]*tmp[i-1,2])/micK)))
    
    tmp[i,2] <-tmp[i-1,2]*(exp(inVec[4]*(1-(tmp[i-1,2]+inVec[2]*tmp[i-1,1])/micK)))
  }
  return(tmp[10,])
}



  for(k in 2:n.gen){

    oneGenPop <- apply(cbind(oneMat[,k-1], twoMat[,k-1]), 1, internalLV)

    nextPop <- apply(oneGenPop, 2, nextSeed, tProb1=inVec[3], tProb2=inVec[4])

    oneMat[,k] <- nextPop[1,]
    twoMat[,k] <- nextPop[2,]

  }

return(cbind(mean(colMeans(oneMat)), mean(colMeans(twoMat))))
}

micMat <- apply(inMat, 1, paramTest, hostMat=hostMat)

tmat <- cbind(inMat, micMat[1,], micMat[2,])

plot(x=1:ncol(micMat), y=seq(min(micMat[1,]), max(micMat[1,]), by=((max(micMat[1,])-min(micMat[,1]))/(ncol(micMat)-1))), type="n", xlab="parameter combination", ylab="microbe 1 population")

points(x=1:ncol(micMat),y=tmat[,6])

```








